Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-21 paper/paper.tex
The paper is almost finished.

Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}. Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} from a text index to a path index. The original GCSA could only index directed acyclic graphs. It depended on pruning the complex regions, as there were no limits on query length. Kim et~al.~\cite{Kim2015-2016} combined GCSA with the HISAT read aligner \cite{Kim2015}. The result was the first practical graph-based read aligner, though it uses the graph for more accurate mapping to a linear reference.

@@

This paper contains several lemmas describing the properties of generalizations of de~Bruijn graphs. Their proofs are included in Appendix~\ref{appendix:proofs}.

@@

\begin{lemma}[Short keys]\label{lemma:pg-keys}
Let $G' = (V', E')$ be a path graph, where $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}$ for all edges $(u', v') \in E'$. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$; and
(b) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

@@

\begin{lemma}[Pruning]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v)$.
If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus V'_{K}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
\end{lemma}

@@

Path graphs are essentially equivalent to the prefix-sorted automata used in the original paper on GCSA \cite{Siren2014}. As many of the technical details are different, we now take a look at how a GCSA of a path graph works. We also extend GCSA with techniques based on suffix trees.
See Figure~\ref{figure:pruned-index} for an example of a GCSA.
Appendix~\ref{appendix:construction} describes a construction algorithm that builds a GCSA for a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph from a set of paths of length $k$. While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.

@@

Let $G' = (V', E')$ be a path graph. We sort the nodes $V'$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$. For each node $v' \in V'$, we append
\begin{itemize}
\item $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$;
\item $\bvIN$ with the indegree $x = G.\gindegree(v')$ encoded as a binary sequence $0^{x-1} 1$; and
\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v')$ encoded as a binary sequence $0^{x-1} 1$.
\end{itemize}
If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of predecessor nodes $u'$ are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v')$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

Consider now the outgoing edges. If the lexicographic rank of node $u' \in V'$ is $i'$, the range of outgoing edges $(u', v') \in E'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$. The edges are already sorted by keys $G'.\gkey(u')$. All outgoing edges from the same node have the same sort key, as there is nothing to distinguish them from one another. Because graph $G'$ is a path graph, we know that key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. The sorting orders are therefore compatible. For every $j \in [sp_{in}, ep_{in}]$ for node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for node $u' \in V'$ implies an edge $(u', v') \in E'$. If we sort the incoming edges $(u', v') \in E'$ stably by strings $G'.\glabel(u') \cdot G'.\gkey(v')$, we have $(u'_{1}, v'_{1}) < (u'_{2}, v'_{2})$ in that order, if $G'.\gkey(u'_{1}) < G'.\gkey(u'_{2})$.

We implement query $G'.\find(X)$ with \emph{backward searching}. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes of the path graph. Now assume that $\abs{X} \ge 1$ and that query $G'.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range corresponding to the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$. We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i}], X[i])$.

Query $G'.\locate(X)$ requires retrieving the values $G'.\gvalue(v')$ for nodes $v' \in V'$ in the lexicographic range returned by query $G'.\find(X)$ and removing duplicates. Storing the values explicitly for all nodes would make the path index several times larger than the graph structure described above. To avoid that, GCSA uses a similar \emph{sampling} scheme to the one used in FM\nobreakdash-indexes. We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are integers, and that the node numbering has been chosen with our sampling scheme in mind. If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.

Let $v' \in V'$ be a node of the path graph. We sample the values $G'.\gvalue(v')$, (a) if there are multiple incoming edges to node $v'$; (b) if $v'$ is the source node $s$; or (c) if $(u', v') \in E'$ is the only incoming edge to node $v'$, and $G'.\gvalue(v') \ne \set{u+1 \mid u \in G'.\gvalue(u')}$. We may also sample the values for some nodes on long unary paths for performance reasons. If the values $G'.\gvalue(v')$ has not been sampled, we can derive them from sampled values by following the incoming edges backwards.

Following edges backwards is based on a similar generalization of \LFmapping{} as backward searching. If node $v' \in V'$ has lexicographic rank $i$, the lexicographic rank of its only predecessor is
$G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$. If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.

Let $B_{S}[0, \abs{V'}-1]$ be a bitvector marking the sampled nodes. If we have sampled the values for the node $v' \in V'$ with lexicographic rank $i$, we mark that as $B_{S} = 1$. We can then determine the rank of node $v'$ among the sampled nodes as $j = B_{S}.\rank(i, 1)$. For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as in bitvectors $\bvIN$ and $\bvOUT$. We then store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each. The sampled values for node $v'$ with rank $j$ among the sampled nodes can be found at $V_{S}[B_{V}.\select(j, 1) + 1, B_{V}.\select(j+1, 1)]$.

If $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$, such as in de~Bruijn graphs and maximally pruned de~Bruijn graphs, we can use a simplified encoding for the graph. We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$. If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$. Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
$$
G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
$$
We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector. On the other hand, computing $G'.\LF(i)$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$. The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.

@@

GCSA is a generalization of FM\nobreakdash-indexes for path graphs. Because path graphs are quite similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them to work with path graphs.

Let $G' = (V', E')$ be an \orderk{k} (pruned) de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys. We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree. The compacted trie can be considered the \emph{suffix tree of the path graph}.

@@

If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern. Hence we can safely use $\parent$ queries with such ranges.

BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference. It stores both the reference and its reverse complement are stored in the same index, which allows it to extend the pattern in both directions. GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains the reverse complement $\revcomp{K}$). Instead of using bidirectional BWT, we search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}. We have implemented a GCSA-based read aligner using that algorithm in vg \cite{Garrison2014-2016}.

In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range. In GCSA, each node $v' \in V'$ may have multiple values (pointers), and each value may occur in multiple nodes. Determining the  number of distinct start nodes with a $\locate$ query can be slow. Hence we need additional structures to support efficient \emph{counting queries}. The problem is similar to determining the \emph{document frequency} of a pattern. Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern. We can augment an FM\nobreakdash-index with a bitvector of length $2n-d$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.

Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$ in the collection, and let $\countq(v)$ be the document frequency of the label $\ell(v)$. If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, we determine the number of \emph{redundant} documents in node $v$ as $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$. Consider now the \emph{inorder traversal} of the suffix tree, where we may encounter the same nodes multiple times. We create an array $R[0, n-2]$ of the values $R(v)$ over all internal nodes $v$. If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$. For subsequent visits to the same node, we set $R[j] = 0$. Range $R[sp, ep-1]$ now covers the internal nodes in the subtree rooted at node $v$. We can now determine document frequencies as $\countq(v) = (ep+1-sp) - \sum_{i=sp}^{ep-1} R[i]$. If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}[0, n-d-1]$, where we can compute the sums with $\select$ queries:
$$
\sum_{i=a}^{b} R[i] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
$$

For value counting, we use array $R$ to store the number of redundant values in each suffix tree node. We also use another array $A[0, \abs{V'}-1]$, which tells the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$. We then encode array $A$ as a bitvector $B_{A}$ in the same way as bitvector $B_{R}$ above. The number of distinct start nodes for paths matching pattern $X$ (with $\abs{X} \le k$) is then
$$
G'.\countq(X) = G'.\countq(G'.\find(X)) = (ep+1-sp) + \sum_{i=sp}^{ep} A[i] - \sum_{i=sp}^{ep-1} R[i],
$$
where $[sp, ep]$ is the lexicographic range returned by the $\find$ query. The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.

@@

GCSA2 has been implemented as the path index library of vg \cite{Garrison2014-2016}. The implementation is in C++ and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). As the implementation uses the \emph{Succinct Data Structures Library 2.0} (SDSL) \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.

The \emph{variation graphs} used in vg have strings as node labels. The average length of a node label is typically around 10~bases. Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples for performance reasons.

@@

We built vg graphs based from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}. To avoid excessive growth, we pruned complex regions of the graphs using \texttt{vg mod -p -l 16 -e 4 | vg mod -S -l 100}, removing paths where \kmer{16}s cross more than $4$ nontrivial edges and subgraphs shorter than $100$ bases. We then extracted paths of length $16$ from the forward strands of all chromosomes. There were a total of 4.80~billion paths with 1.53~billion distinct labels. We then built GCSA2 using $1$, $2$, and $3$ doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes, respectively. The results are summarized in Table~\ref{table:construction}.

@@

A single-strand whole-genome index can be built overnight with 96~gigabytes of memory. Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower. The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph. It consists of a pruning step, followed by merging the paths with identical labels.} The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned before index construction. The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index. There are two main reasons: (1) we use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}; and (2) $\locate$ queries generate duplicates, which must be filtered out. Switching to the simplified encoding for maximally pruned graphs would make $\find$ queries faster and $\locate$ queries slower.

>>>>

Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}. Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) for directed acyclic graphs. It depended on pruning the complex regions, as there were no limits on query length. Kim et~al.~\cite{Kim2015-2016} combined GCSA with the HISAT read aligner \cite{Kim2015}. The result was the first practical graph-based read aligner, though it uses the graph for more accurate mapping to a linear reference.

@@

The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendixes contain more material from the full paper.

@@

\begin{lemma}[Short keys]\label{lemma:pg-keys}
Let $G' = (V', E')$ be a path graph with $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all edges $(u', v') \in E'$. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$; and
(b) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

@@

\begin{lemma}[Pruning]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v' \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v')$.
If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u', v' \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus V'_{K}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
\end{lemma}

@@

We now show how a GCSA of a path graph works. As path graphs are not exactly the same as the prefix-sorted automata used with the original GCSA \cite{Siren2014}, the technical details will be different. We also extend GCSA with techniques based on suffix trees.
See Figure~\ref{figure:pruned-index} for an example of a GCSA index.
Appendix~\ref{appendix:construction} gives a construction algorithm for the GCSA of a pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph, starting from a set of paths of length $k$. While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.

@@

Let $G' = (V', E')$ be a path graph. We sort the nodes $V'$ by their keys in lexicographic order and generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$ from the nodes in that order. For each node $v' \in V'$, we append $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$; $\bvIN$ with the indegree encoded as $0^{G.\gindegree(v')-1} 1$; and $\bvOUT$ with the outdegree as $0^{G.\goutdegree(v')-1} 1$.

If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

The range of outgoing edges $(u', v') \in E'$ for node $u' \in V'$ with lexicographic rank $i'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$. The edges are already sorted by keys $G'.\gkey(u')$. Because graph $G'$ is a path graph, we know that key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. The sorting orders are therefore compatible. For every $j \in [sp_{in}, ep_{in}]$ for a node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for a node $u' \in V'$ implies an edge $(u', v') \in E'$. If we sort the incoming edges $(u', v') \in E'$ stably by strings $G'.\glabel(u') \cdot G'.\gkey(v')$, we have $(u'_{1}, v'_{1}) < (u'_{2}, v'_{2})$ in that order, if $G'.\gkey(u'_{1}) < G'.\gkey(u'_{2})$.

We use \emph{backward searching} for query $G'.\find(X)$. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes. Now assume that $\abs{X} \ge 1$ and that $G'.\find(X[i+1, \abs{X}-1]) = [sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range $G'.\find(X[i, \abs{X}-1])$, which is the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$. We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges; the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i+1}], X[i])$.

Query $G'.\locate(X)$ retrieves the values $G'.\gvalue(v')$ for nodes $v' \in G'.\find(X)$ and filters out duplicates. Storing the values explicitly for all nodes would require too much space. Like FM\nobreakdash-indexes, GCSA uses a \emph{sampling} scheme to avoid that. We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are conveniently chosen integers. If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.

We sample the values $G'.\gvalue(v')$ for a node $v' \in V'$, (a) if there are multiple incoming edges to node $v'$; (b) if $v'$ is the source node $s$; or (c) if $G'.\gvalue(v') \ne \set{u+1 \mid u \in G'.\gvalue(u')}$ for the only incoming edge $(u', v') \in E'$. We may also sample the values for some nodes on long unary paths for performance reasons. If the values $G'.\gvalue(v')$ have not been sampled, we can derive them from sampled values by following the incoming edges.

If node $v' \in V'$ with lexicographic rank $i$ has only one predecessor, the lexicographic rank of the predecessor is $G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$ . If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.

Let $B_{S}[0, \abs{V'}-1]$ be a bitvector. If we have sampled the values for the node $v' \in V'$ with lexicographic rank $i$, we mark that as $B_{S}[i] = 1$. We can then determine the rank of node $v'$ among the sampled nodes as $j = B_{S}.\rank(i, 1)$. For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as in bitvectors $\bvIN$ and $\bvOUT$. We store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each. The sampled values for node $v'$ with rank $j$ among the sampled nodes can be found at $V_{S}[B_{V}.\select(j, 1) + 1, B_{V}.\select(j+1, 1)]$.

@@

GCSA is a generalization of the FM\nobreakdash-index for path graphs. Because path graphs are similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them for path graphs.

Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by the keys. We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree. The compacted trie can be considered the \emph{suffix tree} of graph $G'$.

@@

If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern. Hence we can safely use $\parent$ queries with such ranges.

BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference. It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions. GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains key $\revcomp{K}$). However, we can search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.

In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range. In GCSA, each node $v' \in V'$ may have multiple values (pointers), and a value may occur in multiple nodes. Determining the number of distinct values with a $\locate$ query can be slow. Hence we need additional structures to support efficient \emph{counting queries}. The problem is similar to determining the \emph{document frequency} of a pattern. Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern. We can augment an FM\nobreakdash-index with a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.

Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$, and let $\countq(v)$ be the document frequency of the label $\ell(v)$. If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, the number of \emph{redundant} documents in them is $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$. We create an array $R[0, n-2]$ based on the \emph{inorder traversal} of the suffix tree. If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$. We set $R[j] = 0$ for any subsequent visits to the same node. Range $R[sp, ep-1]$ covers the internal nodes in the subtree with node $v$ as the root. We can determine document frequencies as $\countq(v) = (ep+1-sp) - \sum R[sp, ep-1]$. If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}$, where we can compute the sums with $\select$ queries:
$$
\sum R[a, b] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
$$

For value counting, we use array $R$ to store the number of redundant values as above. We use another array $A[0, \abs{V'}-1]$ to store the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$, and encode it as a bitvector $B_{A}$ in the same way as array $R$ above. The number of distinct values in range $[sp,ep] = G'.\find(X)$ for a pattern $X$ is
$$
G'.\countq(X) = G'.\countq([sp, ep]) = (ep+1-sp) + \sum A[sp, ep] - \sum R[sp, ep-1].
$$
The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.

@@

GCSA2 has been developed as the path index library of vg \cite{Garrison2014-2016}. The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.

The \emph{variation graphs} used in vg have strings as node labels. The average length of a label is typically around 10~bases. Each node can be traversed in both forward and reverse complement orientations, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples for performance reasons.

@@

We built vg graphs from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}. To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial with \texttt{vg mod -p -l -e 4} and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}. We extracted paths of length $16$ from the forward strands of all chromosomes. There were a total of 4.80~billion paths with 1.53~billion distinct labels. We built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes. The results are summarized in Table~\ref{table:construction}.  We could build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache. The de~Bruijn graph had roughly $1.031^{k} \cdot 2.38$~billion nodes, but pruning reduced the size to 4.4\nobreakdash--5.8~billion nodes. The index grew more slowly with $k$, as we sampled the values more often in lower order graphs.

@@

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. The $\find$ and $\locate$ queries are several times slower than in a fast whole-genome FM\nobreakdash-index. We have to use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}, while our $\locate$ queries generate duplicates, which must be filtered out. A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.

\begin{table}[t!]
\begin{center}
\caption{Query benchmarks using an \orderk{64} index. Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
\begin{tabular}{c|cc|cc|cccc}
\hline
\textbf{Length} & \textbf{Patterns} & \textbf{Occs} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
16 & 351584 & 347453 & 1829M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 468 \textmu{}s \\
64 & 351567 & 326101 & 0.33M & 1.35M & 88.9 \textmu{}s & 0.36 \textmu{}s & 0.42 \textmu{}s & 10.3 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}
